# Cross Sectional Multilevel Models

```{r}
#| label: unconditionalsetup
#| echo: false
#| output: false

library(Statamarkdown)

library(lme4) 

library(JuliaCall)

julia_setup(JULIA_HOME = "/Applications/Julia-1.8.app/Contents/Resources/julia/bin")

```

## The Equation

Recall the general model of @eq-MLMsimple, and the syntax outlined in @sec-syntax. Below in @eq-MLMsubstantive, we consider a more substantive example.

$$\text{outcome}_{ij}= \beta_0 + \beta_1 \text{warmth}_{ij} +$$ {#eq-MLMsubstantive}

$$\beta_2 \text{physical punishment}_{ij} +$$

$$\beta_3 \text{identity}_{ij} + \beta_4 \text{intervention}_{ij} + \beta_5 \text{HDI}_{ij} +$$

$$u_{0j} + u_{1j} \times \text{warmth}_{ij} + e_{ij}$$ 

## Correlated and Uncorrelated Random Effects

Consider the covariance matrix of random effects (e.g. $u_{0j}$ and $u_{1j}$). In @eq-varcovar the covariances of the random effects are constrained to be zero.

$$\begin{bmatrix}
var(u_{0j}) & 0 \\
0 & var(u_{1j}) 
\end{bmatrix}$$ {#eq-varcovar}

As discussed in the Chapter on multilevel models with cross-sectional data, however, one can consider a multilevel model in which the random effects are correlated, as is the case in @eq-varcovaruns.

$$\begin{bmatrix}
var(u_{0j}) & cov(u_{0j}, u_{1j}) \\
cov(u_{0j}, u_{1j}) & var(u_{1j}) 
\end{bmatrix}$$ {#eq-varcovaruns}

Procedures for estimating models with uncorrelated and correlated random effects are detailed below [@StataCorp2021:2; @JSSv067i01; @MixedModels].

-------------------------------------------------------------
Software    Uncorrelated          Correlated
            Random                Random
            Effects               Effects
----------- --------------------- ---------------------------
Stata       default               add option: `, cov(uns)`

R           separate random       separate random 
            effects from          effects from
            grouping variable     grouping variable
            with `||`             with `|`

Julia       separate terms        separate random 
            for each random       effects from
            effect e.g.           grouping variable
            `(1 | group) +`       with `|`.
            `(0 + x | group)`
-------------------------------------------------------------

: Correlated and Uncorrelated Random Effects {#tbl-REs}

All models in the examples below are run with *uncorrelated* random effects, but could just as easily be run with *correlated* random effects. 

## Run Models

::: {.panel-tabset group="language"}

### Stata

#### Get The Data 

```{stata, collectcode=TRUE}

use simulated_multilevel_data.dta

```

#### Run The Model

```{stata}
#| label: Stata-cross-sectional

mixed outcome warmth physical_punishment i.identity i.intervention HDI || country: warmth

```

### R

#### Get The Data

```{r}

library(haven)

df <- read_dta("simulated_multilevel_data.dta")

```

#### Run The Model

::: {.callout-caution collapse="false"}
`lme4` does not directly provide p values in results, because of some disagreement over exactly how these p values should be calculated. Therefore, in this Appendix, I also call library `lmerTest` to provide p values for `lme4` results.
:::

```{r}
#| warning: false
#| label: R-cross-sectional

library(lme4) 

library(lmerTest)

fit1 <- lmer(outcome ~ warmth + physical_punishment + 
               identity + intervention + HDI +
               (1 + warmth || country),
             data = df)

summary(fit1)

```

### Julia

#### Get The Data

```{julia}
#| output: false
#| label: Julia-cross-sectional

using Tables, MixedModels, StatFiles, DataFrames, CategoricalArrays, DataFramesMeta

df = DataFrame(load("simulated_multilevel_data.dta"))

```

#### Change Country To Categorical

```{julia}
#| output: false

@transform!(df, :country = categorical(:country))

```

#### Run The Model

```{julia}

m1 = fit(MixedModel, @formula(outcome ~ warmth + physical_punishment + 
               identity + intervention + HDI +
               (1 | country) +
               (0 + warmth | country)), df)

```


:::

## Interpretation

Models suggest that parental warmth is associated with increases in the beneficial outcome, while physical punishment is associated with decreases in the beneficial outcome. Membership in the group represented by `identity` is not associated with the outcome. The intervention is associated with increases in the outcome. The Human Development Index is not associated with the outcome.





