# Descriptive Statistics

```{r}
#| label: unconditionalsetup
#| echo: false
#| output: false

library(Statamarkdown)

library(lme4) 

library(JuliaCall)

julia_setup(JULIA_HOME = "/Applications/Julia-1.8.app/Contents/Resources/julia/bin")

```

## Descriptive Statistics

::: {.panel-tabset group="language"}

### Stata

```{stata, collectcode=TRUE}

use simulated_multilevel_data.dta // use data

```

We use `summarize` for *continuous* variables, and `tabulate` for *categorical* variables. 

```{stata, collectcode=TRUE}
#| label: stata-descriptives

summarize outcome warmth physical_punishment HDI

tabulate identity

tabulate intervention

```

### R 

```{r}

library(haven) # read data in Stata format

df <- read_dta("simulated_multilevel_data.dta")

```

R's descriptive statistics functions rely heavily on whether a variable is a *numeric* variable, or a *factor* variable. Below, I convert two variables to factors (`factor`) before using `summary`[^skimr] to generate descriptive statistics.

[^skimr]: `skimr` is an excellent new alternative library for generating descriptive statistics in R.

```{r}
#| label: R-descriptives

df$country <- factor(df$country)

df$identity <- factor(df$identity)

df$intervention <- factor(df$intervention)

summary(df)

```

### Julia

```{julia}
#| output: false
#| label: Julia-descriptives

using Tables, MixedModels, MixedModelsExtras, StatFiles, DataFrames, CategoricalArrays, DataFramesMeta

df = DataFrame(load("simulated_multilevel_data.dta"))

```

Similarly to R, Julia relies on the idea of *variable type*. I use `transform` to convert the appropriate variables to *categorical* variables. 

```{julia}
#| output: false

@transform!(df, :country = categorical(:country))

@transform!(df, :identity = categorical(:identity))

@transform!(df, :intervention = categorical(:intervention))

```

```{julia}

describe(df) # descriptive statistics

```

:::

## Interpretation

Examining descriptive statistics is an important first step in any analysis. It is important to examine your descriptive statistics first, before skipping ahead to more sophisticated analyses, such as multilevel models.

In examining the descriptive statistics for this data, we get a sense of the data.

* `outcome` has a mean of approximately 52 and ranges from approximately 30 to 75.
* `warmth` and `physical punishment` are both variables that represent the number of times that parents use each of these forms of discipline in a week. The average of the former is about 3.5, while the average of the latter is about 2.5.
* `HDI`, the Human Development Index has an average of about 65, and a wide range.
* `identity` is a categorical variable for a hypothetical identity group, and has values of 0 and 1.
* `intervention` is also a categorical variable, and has values of 0 and 1.







