# Multilevel Multilingual

## Introduction

Below, I describe the use of [Stata](https://www.stata.com/), [R](https://www.r-project.org/), and [Julia](https://www.julialang.org/) to estimate multilevel models. 
## The Data

The examples use the `simulated_multilevel_data.dta` file from [*Multilevel Thinking*](https://agrogan1.github.io/multilevel-thinking/simulated-multi-country-data.html). Here is a [direct link](https://github.com/agrogan1/multilevel-multilingual/raw/main/simulated_multilevel_data.dta) to download the data.

```{r}
#| echo: false
#| output: false

file.copy("../multilevel-thinking/simulate-and-analyze-multilevel-data/simulated_multilevel_data.dta", 
          ".",
          overwrite = TRUE) # copy over most recent version

```

```{r}
#| echo: false
#| label: tbl-multilingual1
#| tbl-cap: Sample of Simulated Multilevel Data

library(haven) # read Stata

library(pander) # nice tables

df <- read_dta("simulated_multilevel_data.dta") # read in data

pander(head(df)) # nice table

```

```{r}
#| echo: false
#| output: false

# KEEP THIS CHUNK FOR INFORMATIONAL PURPOSES!!!

library(JuliaCall)

# OS dependent location of Julia

switch(Sys.info()[['sysname']],
       Windows= {julia_setup(JULIA_HOME = "C:/Users/agrogan/AppData/Local/Programs/Julia-1.10.2/bin/")},
       Darwin = {julia_setup(JULIA_HOME = "/Applications/Julia-1.8.app/Contents/Resources/julia/bin")})

```

## An Introduction To Equations and Syntax

To explain statistical syntax for each software, I consider the general case of a multilevel model with dependent variable `y`, independent variables `x` and `z`, clustering variable `group`, and a random slope for `x`. *i* is the index for the person, while *j* is the index for the `group`.

$$y = \beta_0 + \beta_1 x_{ij} + \beta_2 z_{ij} + u_{0j} + u_{1j} \times x_{ij} + e_{ij}$$ {#eq-MLMsimple}

::: {.panel-tabset group="language"}

### Stata

In Stata `mixed`, the syntax for a multilevel model of the form described in @eq-MLMsimple is:

```{stata}
#| echo: true
#| eval: false

mixed y x || group: x

```

### R

In R `lme4`, the general syntax for a multilevel model of the form described in @eq-MLMsimple is:

```{r}
#| echo: true
#| eval: false

lmer(y ~ x + z + (1 + x || group), data = ...)

```

### Julia

In Julia `MixedModels`, the general syntax for a multilevel model of the form described in @eq-MLMsimple is:

```{julia}
#| echo: true
#| eval: false

fit(MixedModel, @formula(y ~ x + z + (1 + x | group)), data)

```

:::

