# Longitudinal Multilevel Models {#sec-longitudinal}

```{r}
#| label: longitudinalsetup
#| echo: false
#| output: false

library(Statamarkdown)

library(lme4) 

library(JuliaCall)

julia_setup(JULIA_HOME = "/Applications/Julia-1.10.app/Contents/Resources/julia/bin")

```

## The Data

The data employed in these examples are a longitudinal extension of the data described in @sec-data.

## The Equation

$$\text{outcome}_{itj} = \beta_0 + \beta_1 \text{parental warmth}_{itj} + \beta_2 \text{physical punishment}_{itj} + \beta_3 \text{time}_{itj} \ + $$ {#eq-MLM-longitudinal} 

$$\beta_4 \text{identity}_{itj} + \beta_5 \text{intervention}_{itj} + \beta_6 \text{HDI}_{j} +$$

$$u_{0j} + u_{1j} \times \text{parental warmth}_{itj} \ + $$

$$v_{0ij} + v_{1ij} \times \text{time}_{itj} + e_{itj}$$

## Growth Trajectories

Remember, following the discussion in the main text, that in longitudinal multilevel models, the variable for *time* assumes an important role as we are often thinking of a *growth trajectory* over time.

As discussed in the main text, think about a model where *identity* is a (1/0) variable for membership in one of two groups:

$$\text{outcome} = \beta_0 + \beta_t \text{time} + \beta_\text{identity} \text{identity} + \beta_\text{interaction} \text{identity} \times \text{time} + u_{0i} + e_{it}$$

Then, each identity group has its own intercept and time trajectory:

| Group | Intercept                         |  Slope (Time Trajectory)             |
|-------|-----------------------------------|--------------------------------------|
| 0     | $\beta_0$                         | $\beta_t$                            |
| 1     | $\beta_0 + \beta_\text{identity}$ | $\beta_t + \beta_\text{interaction}$ |

: Slope and Intercept for Each Group {#tbl-trajectory}


::: {.callout-tip}
## Main Effects and Interactions

Thus, again following the main text, in longitudinal multilevel models, *main effects* modify the *intercept* of the time trajectory, while *interactions with time*, modify the *slope* of the time trajectory. Below, we run models with *main effects only*, then models with *main effects, and interactions with time*.
:::

## Run Models

::: {.callout-warning}
Remember that we are estimating a model in which time points are nested inside families, who are in turn nested inside countries. For each software package, it is accordingly important to specify the way in which different levels of the data are nested. Pay careful attention to the syntax examples below with regard to `id` and `country` 
:::

::: {.panel-tabset group="language"}

### Stata

#### Get The Data 

```{stata, collectcode=TRUE}

use simulated_multilevel_longitudinal_data.dta

```

#### Run The Models

##### Main Effects Only

```{stata}
#| label: Stata-maineffects

mixed outcome t warmth physical_punishment i.identity i.intervention HDI || /// 
country: warmth || id: t

```

##### Interactions With Time

```{stata}
#| label: Stata-interactions

mixed outcome c.t##(c.warmth c.physical_punishment i.identity i.intervention c.HDI) || country: warmth || id: t

```

### R

#### Get The Data

```{r}

library(haven)

dfL <- read_dta("simulated_multilevel_longitudinal_data.dta")

```

#### Change Some Variables To Categorical

```{r}

dfL$identity <- factor(dfL$identity)

dfL$intervention <- factor(dfL$intervention)

```

#### Run The Models

::: {.callout-caution collapse="false"}
`lme4` does not directly provide p values in results, because of some disagreement over exactly how these p values should be calculated. Therefore, in this Appendix, I also call library `lmerTest` to provide p values for `lme4` results.
:::

::: {.callout-tip}
R prefers to use scientific notation when possible. I find that the use of scientific notation can be confusing in reading results. I turn off scientific notation by setting a penalty for its use:  `options(scipen = 999)`.  
:::

##### Main Effects Only

```{r}
#| label: R-maineffects

library(lme4) 

library(lmerTest)

options(scipen = 999) 

fit2A <- lmer(outcome ~ t + warmth + physical_punishment + 
               identity + intervention + HDI +
               (1 | country/id),
             data = dfL)

summary(fit2A)

```

##### Interactions With Time

```{r}
#| label: R-interactions

fit2B <- lmer(outcome ~ t *(warmth + physical_punishment + 
               identity + intervention + HDI) +
               (1 | country/id),
             data = dfL)

summary(fit2B)

```


### Julia

#### Get The Data

```{julia}
#| output: false
#| label: Julia-getdata

using Tables, MixedModels, StatFiles, DataFrames, CategoricalArrays, DataFramesMeta

dfL = DataFrame(load("simulated_multilevel_longitudinal_data.dta"))

```

#### Change Some Variables To Categorical

```{julia}
#| output: false

@transform!(dfL, :country = categorical(:country))

@transform!(dfL, :identity = categorical(:identity))

@transform!(dfL, :intervention = categorical(:intervention))

```

#### Run The Models

##### Main Effects Only

```{julia}
#| label: Julia-maineffects

m2A = fit(MixedModel, @formula(outcome ~ t + warmth + 
                                 physical_punishment + 
                                 identity + intervention + 
                                 HDI +
                                 (1 | country) + 
                                 (0 + warmth | country) +
                                 (1 | id)), dfL)

```

##### Interactions With Time

```{julia}
#| label: Julia-interactions

m2B = fit(MixedModel, @formula(outcome ~ t * (warmth + 
                                 physical_punishment + 
                                 identity + intervention + 
                                   HDI) +
                                 (1 | country) +
                                 (0 + warmth | country) +
                                 (1 | id)), dfL)

```

:::

## Interpretation

The *main effects only model* suggests that time is associated with increases in the outcome. In the main effects model, main effects other than time, indicate whether a particular variable is associated with higher or lower intercepts of the time trajectory, at the beginning of the study time. Warmth is again associated with increases in the outcome, while physical punishment is associated with decreases in the outcome. Identity is again not associated with the outcome, while the intervention is associated with higher levels of the outcome. The Human Development Index is again not associated with the outcome.

The second model adds interactions with time to the first model. Results are largely similar to the prior model. However, here we not only examine whether main effects other than time are associated with higher or lower time trajectories, but also whether particular variables are associated with differences in the slope of the time trajectory. In this case, we find insufficient evidence that any independent variable is associated with changes in the slope of the time trajectory.

::: {.callout-tip}
## Which Interactions To Test?

In this example--for the sake of illustration--I test the interaction of *every* independent variable with time. In many cases, it may make sense to test only only one or two interactions of time with particular variables of key interest. Also, after finding, as I did in this model, that none of the interactions of other independent variables with time are significant, I might report the model with interactions, or might report only the results of the model with only main effects. 
:::

It may be illustrative to imagine how we would interpret the results had a particular interaction term been statistically significant. Let us consider one of the interaction terms with the largest coefficient, `intervention#time`. The interaction of the intervention with time is positive. Had this coefficient been statistically signifcant, it would have indicated that the intervention was associated with more rapid increases in the outcome over time *in addition to* the fact that the intervention is associated with higher initial levels of the outcome.








