---
title: "# Models With Three or More Levels and Cross-Classified Models"
date: "today"
format:
  pdf: 
    toc: true
    toc-depth: 5
  html:
    toc: true
bibliography: multilevel-multilingual.bib
---

```{r}
#| label: morelevelssetup
#| echo: false
#| output: false

library(Statamarkdown)

library(lme4) 

library(JuliaCall)

julia_setup(JULIA_HOME = "/Applications/Julia-1.10.app/Contents/Resources/julia/bin")

```

## Introduction

A two level multilevel model imagines that *Level 1* units are nested in *Level 2* units. A three level multilevel model imagines that *Level 1* units are nested in *Level 2* units, which are in turn nested in *Level 3*. As more levels are added to the model (e.g. *Level 4*), we imagine all of these levels to be hierarchically nested.

A *cross classified* model imagines that the nesting is not hierarchical, but rather that there are two sets of clusters or nestings in which individuals may be nested. 

Below, I describe the use of [Stata](https://www.stata.com/), [R](https://www.r-project.org/), and [Julia](https://www.julialang.org/) to estimate these models.

## Three Or More Levels {#sec-fourlevel}

### The Data

I use the *longitudinal* data from *Multilevel Thinking* to which I have added an extra level of *United Nations Region* [@ArelBundock2018]. This data thus requires a four level model.

### The Equation

$$\text{outcome}_{itjk} = \beta_0 + \beta_1 \text{parental warmth}_{itjk} + \beta_2 \text{physical punishment}_{itjk} + \beta_3 \text{time}_{itjk} \ + $$ {#eq-MLM-fourlevel-multilingual} 

$$\beta_4 \text{identity}_{itjk} + \beta_5 \text{intervention}_{itjk} + \beta_6 \text{HDI}_{itjk} +$$

$$w_{0k} + u_{0j} + v_{0i} + e_{itjk}$$ 

Here we imagine $w_{0k}$ (region), $u_{0j}$ (country) and $v_{0i}$ (family) are hierarchically nested effects. 


### Run The Models

::: {.panel-tabset group="language"}

#### Stata

##### Get The Data

```{stata, collectcode=TRUE}

use "fourlevel.dta", clear

```

##### Unconditional Model

```{stata, collectcode=TRUE}

mixed outcome || UNregion: || country: || family:

```

##### Conditional Model

```{stata, collectcode=TRUE}

mixed outcome t warmth physical_punishment i.identity i.intervention HDI ///
|| UNregion: || country: || id:
  
```

#### R

##### Get The Data

```{r}

library(haven)

df4 <- read_dta("fourlevel.dta")

```

##### Change Some Variables To Categorical

```{r}

df4$identity <- factor(df4$identity)

df4$intervention <- factor(df4$intervention)

```

##### Unconditional Model

::: {.callout-caution collapse="false"}
`lme4` does not directly provide p values in results, because of some disagreement over exactly how these p values should be calculated. Therefore, in this Appendix, I also call library `lmerTest` to provide p values for `lme4` results.
:::

::: {.callout-tip}
R prefers to use scientific notation when possible. I find that the use of scientific notation can be confusing in reading results. I turn off scientific notation by setting a penalty for its use:  `options(scipen = 999)`.  
:::

```{r}

library(lme4) 

library(lmerTest)

options(scipen = 999) 

fit4A <- lmer(outcome ~  (1 | UNregion/country/id),
             data = df4)

summary(fit4A)

```

##### Conditional Model

```{r}

fit4B <- lmer(outcome ~ t + warmth + physical_punishment + 
                identity + intervention + HDI + 
                (1 | UNregion/country/id),
              data = df4)

summary(fit4B)

```

#### Julia

##### Get The Data

```{julia}
#| output: false
#| label: Julia-getdata
#| eval: true

using Tables, MixedModels, StatFiles, DataFrames, CategoricalArrays, DataFramesMeta

df4 = DataFrame(load("fourlevel.dta"))

```

##### Change Some Variables To Categorical

```{julia}
#| output: false
#| eval: false

@transform!(df4, :country = categorical(:country))

@transform!(df4, :UNregion = categorical(:UNregion))

@transform!(df4, :identity = categorical(:identity))

@transform!(df4, :intervention = categorical(:intervention))

```

##### Unconditional Model

```{julia}
#| label: Julia-fourlevelA
#| eval: true

m4A = fit(MixedModel, @formula(outcome ~ t + warmth + 
                                  physical_punishment + 
                                  identity + intervention + 
                                  HDI +
                                  (1 | UNregion) + 
                                  (1 | country) + 
                                  (1 | id)), df4)

```

##### Conditional Model

```{julia}
#| label: Julia-fourlevelB
#| eval: true

m4B = fit(MixedModel, @formula(outcome ~ t + warmth + 
                                  physical_punishment + 
                                  identity + intervention + 
                                  HDI +
                                  (1 | UNregion) + 
                                  (1 | country) + 
                                  (1 | id)), df4)

```

:::

### Interpretation

There is group level variation attributable to individual, country, and region.

As in other models, parental warmth, and participation in the intervention are associated with increases in the outcome. Parental use of physical punishment is associated with decreases in the outcome.

## Cross-Classified Models

### The Data

I use the *cross-sectional* data from *Multilevel Thinking* to which I have added an extra level of a hypothetical language. 

### The Equation

$$\text{outcome}_{ijm} = \beta_0 + \beta_1 \text{parental warmth}_{ijm} + \beta_2 \text{physical punishment}_{ijm} + \beta_3 \text{time}_{ijm} \ + $$ {#eq-crossclassified-multilingual}

$$\beta_4 \text{identity}_{ijm} + \beta_5 \text{intervention}_{ijm} + \beta_6 \text{HDI}_{ijm} +$$ 

$$u_{0j} + m_{0m} + e_{ijm}$$ 

Here $u_{0j}$ (country) and $m_{0m}$ (language) are not nested hierarchically, but are *cross classified*.

### Run The Models

::: {.panel-tabset group="language"}

#### Stata

##### Get The Data

```{stata, collectcode=TRUE}

use "crossclassified.dta", clear

```

##### Unconditional Model

```{stata}

mixed outcome || _all: R.country || _all: R.language

```

##### Conditional Model

```{stata}

mixed outcome warmth physical_punishment i.identity i.intervention HDI || _all: R.country || _all: R.language

```

#### R

##### Get The Data

```{r}

library(haven)

dfCC <- read_dta("crossclassified.dta")

```

##### Change Some Variables To Categorical

```{r}

dfCC$identity <- factor(dfCC$identity)

dfCC$intervention <- factor(dfCC$intervention)

```

##### Unconditional Model

```{r}

library(lme4) 

library(lmerTest)

options(scipen = 999) 

fitCC_A <- lmer(outcome ~  
                 (1 | country) +
                 (1 | language),
             data = dfCC)

summary(fitCC_A)

```

##### Conditional Model

```{r}

fitCC_B <- lmer(outcome ~ t + warmth + physical_punishment + 
                identity + intervention + HDI + 
                (1 | country) +
                (1 | language),
              data = dfCC)

summary(fitCC_B)

```

#### Julia

##### Get The Data

```{julia}
#| output: false
#| label: Julia-CC-getdata

using Tables, MixedModels, StatFiles, DataFrames, CategoricalArrays, DataFramesMeta

dfCC = DataFrame(load("crossclassified.dta"))

```

##### Change Some Variables To Categorical

```{julia}
#| output: false
#| eval: false

@transform!(dfCC, :country = categorical(:country))

@transform!(dfCC, :language = categorical(:language))

@transform!(dfCC, :identity = categorical(:identity))

@transform!(dfCC, :intervention = categorical(:intervention))

```

##### Unconditional Model

```{julia}
#| label: Julia-CC-unconditional

mCCA = fit(MixedModel, @formula(outcome ~ 
                                 (1 | country) + 
                                 (1 | language)), dfCC)

```

##### Conditional Model

```{julia}
#| label: Julia-CC-conditional

mCCA = fit(MixedModel, @formula(outcome ~ warmth + 
                                 physical_punishment + 
                                 identity + intervention + 
                                 HDI +
                                 (1 | country) + 
                                 (1 | language)), dfCC)

```

:::

### Interpretation

There is group level variation attributable to both language and country.

As in other models, parental warmth, and participation in the intervention are associated with increases in the outcome. Parental use of physical punishment is associated with decreases in the outcome.

