# Unconditional Model

An *unconditional* multilevel model is a model with no independent variables. One should always run an unconditional model as the first step of a multilevel model in order to get a sense of the way that variation is apportioned in the model across the different levels.

## The Equation

$$\text{outcome}_{ij}= \beta_0 + u_{0j} + e_{ij}$$ {#eq-MLMunconditional}

The Intraclass Correlation Coefficient (ICC) is given by:

$$\text{ICC} = \frac{var(u_{0j})}{var(u_{0j}) + var(e_{ij})}$$ {#eq-ICCunconditional}

In a two level multilevel model, the ICC provides a measure of the amount of variation attributable to Level 2.

```{r}
#| label: unconditionalsetup
#| echo: false
#| output: false

library(Statamarkdown)

library(lme4) 

library(JuliaCall)

julia_setup(JULIA_HOME = "/Applications/Julia-1.8.app/Contents/Resources/julia/bin")

```

## Run Models

::: {.panel-tabset group="language"}

### Stata

```{stata, collectcode=TRUE}

use simulated_multilevel_data.dta // use data

```

```{stata, collectcode=TRUE}

mixed outcome || country: // unconditional model
  
```

```{stata}
  
estat icc // ICC

```

### R

```{r}

library(haven)

df <- read_dta("simulated_multilevel_data.dta")

```

```{r}
#| warning: false

library(lme4) # estimate multilevel models

fit0 <- lmer(outcome ~ (1 | country),
             data = df) # unconditional model

summary(fit0)

```

```{r}

library(performance)

performance::icc(fit0) # ICC

```

### Julia

```{julia}
#| output: false

using Tables, MixedModels, MixedModelsExtras, 
StatFiles, DataFrames, CategoricalArrays, DataFramesMeta

df = DataFrame(load("simulated_multilevel_data.dta"))

```

```{julia}
#| output: false

@transform!(df, :country = categorical(:country))

```

```{julia}

m0 = fit(MixedModel, 
         @formula(outcome ~ (1 | country)), df) # unconditional model

```

```{julia}

icc(m0) # ICC

```

:::

## Interpretation

In each case, the software finds that nearly 8% of the variation in the outcome is explainable by the clustering of the observations in each country.





